### 2019-10-27 report/overview.tex
Rework some sentences and notations

Optimal transport aims to define a geometry and a distance on the set of
mesures of probability on a space.

@@

@@

\paragraph{Monge}

We work on a space of probability $\Omega$. We consider two mesures of
probability $\mu$ and $\nu$ on $\mathcal{P}(\Omega)$. We also consider a cost function
$c : \Omega \times \Omega \rightarrow \R$. This function can for instance be the Euclidean distance. Monge's problem is therefore the following

$$ \int_{\Omega} c \left( x, T(x) \right) \mu(x) $$
\vspace{-2mm}
$$ \text{s.t.} \quad \forall A \in \mathcal{P}(\Omega), \, \nu(A) = \mu \left( T^{-1}(A) \right) $$

The function $T$ is called a transport map. It carries the information
regarding the displacement of each elementary unit of probability: the mass
located in $x$ is moved toward $T(x)$. The function $c$ thus allows to
quantify the cost of the displacement of one elementary unit of mass. We
want to minimize the average distance travelled by mass units during the
optimal transport.

@@

The issue with this formulation of an optimal transport problem is that it
doesn't handle discrete probability mesures. Indeed, if $\mu$ has a Dirac
impulse bigger than every Dirac impulse of $\nu$, then the transport map
cannot be build.

@@

@@

\paragraph{Change of transport plan}
The idea of Kantorovich is to change the function $T : \Omega \rightarrow
\Omega$ into a probability $P$ on the product space $\mathcal{P}(\Omega \times
\Omega)$. On the upper example, Monge's formulation forbids to separate the
masse of size 150 into two separate masses. With Kantorovich's formulation,
this operation is accessible: the mass of $150$ is splitted into a mass of
$90$ and a mass of $60$, both sent to different places. In the discrete
case, we can represent what happens with a matrix.

@@

If we note $D$ our disance matrix, we obtain the following formulation in the discrete case:

@@

@@

In the continuous setting, we introduce the following set of joint probabilities:

$$ \Pi(\mu, \nu) = \left\{ P \in \mathcal{P}(\Omega \times \Omega)~|~\forall A, B \in \Omega, \, P(A \times \Omega) = \mu(A), \, P(\Omega \times B) = \nu(B) \right\} $$

This set encodes the conditions of the discrete case but in the very
continuous setting. We consider once again our cost function $c$ defined
in the Monge's problem. Kantorovich's formulation now becomes the following
optimization problem:

$$ \inf_{P \in \Pi(\mu, \nu)} \E_P \left[ c(X, Y) \right] = \int \int c(x, y) P(dx, dy) $$
\subs{Wasserstein distance}

This optimization problem allows us to define a distance between probability mesures:

\DEF{
    Given a probability space $\Omega$, a cost function $c : \Omega \times
    \Omega \rightarrow \R$ and two probabilitiy mesures $\mu$ and $\nu$ in
    $\mathcal{P}(\Omega)$, the \textbf{p-Wasserstein-distance} between $\mu$
    and $\nu$ is defined by :
	$$ W_p(\mu, \nu) = \left( \inf_{P \in \Pi(\mu, \nu)} \int \int c(x,
	y)^p P(dx, dy) \right)^{1 / p} $$
	\vspace{-3mm}
}

This distance allows to compute barycenters between several probability
distributions, and obtain results that are way more natural than barycenters
obtained with other norms, like $l_2$ for instance. The following
figure compares an interpolation (progressive weighted barycenter of two
distributions) obtained via Wasserstein-distance and a linear interpolation.

@@

We recall that the computation of the barycenter $\mu$ of the $N$ distributions
$\nu_i$ associated to the weights $\lambda_i$ boils down to the following
minimization problem.
$$ \min_{\mu \in \mathcal{P}(\Omega)} \sum_{i=1}^N \lambda_i W_p^p(\mu, \nu_i) $$
Now you can observe what the barycenter looks like with $4$ distributions in the \figurename~\ref{shp1}.
\begin{figure}[h]
	\centering
	\includegraphics[scale=0.4]{wasserstein_bar.png}
	\captionsetup{justification=centering}
	\caption{Each image contains several barycenters of the four shapes that are in the corners}
	\label{shp1}
\end{figure}

@@

@@

To do the computations, we embrace the setting of discrete spaces. Our
probability mesures are sum of diracs impulses:

$$ \mu = \sum_{i=1}^n a_i \delta_{x_i} \qquad \nu = \sum_{j =1}^m b_j \delta_{y_j} $$
Our cost function is a matrix $D \in \R_+^{n \times m}$ such that $D_{ij} = d(x_i, y_j)^p$. Our joint distribution $P$ is also a matrix of $\R_+^{n \times m}$ that has to satisfy several constraints that can be represented matricially. The set which $P$ belongs to is:

$$ U(a, b) = \left\{ P \in \R_+^{n \times m}~|~P \mathbbm{1}_m = a, \, P^\trans \mathbbm{1}_n = b \right\} $$

@@

\paragraph{Dual} We now focus on the dual problem. For this we introduce the lagrangian.

$$ L(P, \alpha, \beta) = \langle P, D \rangle + \alpha^\trans (a - P \mathbbm{1}_m) + \beta^\trans (b - P^\trans \mathbbm{1}_n) $$
The objective function of the dual is:
$$ g(\alpha, \beta) = \min_{P \geqslant 0} L(P, \alpha, \beta) = \min_{P \geqslant 0} \alpha^\trans a + \beta^\trans b + \langle P, D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle $$
If the matrix $D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans$ has a negative coefficient, then by having this coefficient grow toward $+\infty$ we obtain a minimum of $-\infty$. But if all the coefficients are non negative, then  $\langle P, D - \alpha
\mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle$ is non negative and we just have to take $P=0$ to obtain a non zero value. We therefore obtain:
$$ g(\alpha, \beta) = \left\{ \begin{array}{ll}
\alpha^\trans a + \beta^\trans b & \text{si } D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \geqslant 0 \\
-\infty & \text{sinon}
\end{array}\right. $$

@@

Using a minimum flow solver, we can solve this linear optimization problem in
$\mathcal{O}(n^3 \log (n))$. But the solution is not stable, as illustrated
on the following figure. If we slightly move our distance $D$ or if we change
the source and destination mesures, the value of $P$ can drastically change,
since $P$ is a vertex of our simplex. One solution is to regularize the problem to fall back on a set $U_\alpha(a, b)$
strictly convex. This set is strictly contained in $U(a, b)$ and the optimal solution of this new problem won't be optimal for the initial problem, but will remain valid.
Finally, the regularization that we will describe allows to obtain an
interative algorithm of small complexity.
\begin{center}
	\begin{tikzpicture}[scale=20, >={latex}]
		\clip (0.2, -0.01) rectangle (0.51, 0.2);
		\draw[blue, thick] (0.1420, 0.1407) -- (0.5970, -0.1407);
		\draw[black, ->, thick] (0.2038, 0.1216) -- (0.2507, 0.1975);
		\fill[brown!40] (0.4503, 0.1444) -- (0.2251, 0.1444) -- (0.3695, 0.0000) -- (0.4503, -0.0000);
		\fill[greenTikz!60] (0.3827, 0.0946) -- (0.3827, 0.1104) -- (0.3804, 0.1148) -- (0.3759, 0.1206) -- (0.3692, 0.1263) -- (0.3602, 0.1307) -- (0.3556, 0.1321) -- (0.3489, 0.1336) -- (0.3264, 0.1336) -- (0.3219, 0.1321) -- (0.3151, 0.1278) -- (0.3129, 0.1234) -- (0.3129, 0.1191) -- (0.3151, 0.1133) -- (0.3241, 0.1018) -- (0.3444, 0.0888) -- (0.3534, 0.0859) -- (0.3714, 0.0859) -- (0.3804, 0.0917);
		\node[blue] at (0.3444, 0.0888) {$\bullet$};
		\node[blue, below=2] at (0.3444, 0.0888) {$P_\alpha$};
		\node[greenTikz] at (0.3489, 0.1119) {$\bullet$};
		\node[red] at (0.3695, 0.0000) {$\bullet$};
		\node[red, above=2] at (0.3695, 0) {$P$};
		\node at (0.482, 0.08) {$U(a, b)$};
		\node at (0.413, 0.12) {$U_\alpha(a, b)$};
		\node at (0.213, 0.165) {$D$};
	\end{tikzpicture}
\end{center}

\subs{Regularization}

The regularization that we choose consist in maximizing the entropy of the
joint-probability $P$. We exhibit a new parameter $\gamma$ that controls
the regularization and we obtain the following distance:

@@

\paragraph{Remark}
Recall that we can consider two random variables $X$ and $Y$ such that $X \sim \mu$, $Y \sim \nu$ and $X,Y \sim P$. The entropy of $P$ is maximum when $X$ and $Y$ are independant, \emph{ie.} when $P = ab^\trans$ and this maximum entropy is thus $H(\mu) + H(\nu)$. 
The regularization therefore gives a matrix whose positive values are more
distributed.

@@

By strict convexity, there exist a unique matrix $P_\gamma$ that minimizes the distance:
$$ P_\gamma = \argmin_{P \in U(a, b)} \langle P, D \rangle - \gamma H(P) $$

@@

We write the new laplacian:

$$ L(P, \alpha, \beta) = \sum_{ij} \left( P_{ij} D_{ij} + \gamma P_{ij} \left( \log_2 P_{ij} - 1 \right) \right) + \alpha^\trans (a - P \mathbbm{1}_m) + \beta^\trans (b - P^\trans \mathbbm{1}_n) $$
We compute the partial derivative with respect to $P_{ij}$:
$$ \dfrac{\partial L}{\partial P_{ij}} = M_{ij} + \gamma \log_2 P_{ij} - \alpha_i - \beta_j $$
The dual is the maximization of the minimum of the laplacian with respect
to $P$. We must look for $P$ such that the partial derivative is equal to zero:
$$ \dfrac{\partial L}{\partial P_{ij}} = 0 \Rightarrow P_{ij} = e^{\frac{\alpha_i}{\gamma}} e^{-\frac{D_{ij}}{\gamma}}  e^{\frac{\beta_j}{\gamma}} = u_i K_{ij} v_j $$
\findem

\paragraph{Sinkhorn}
We have at our disposition an interative method to find this matrix $P_\gamma$. We use the condition $P_\gamma \in U(a,b)$:
$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
\text{diag}(u) K \text{diag}(v) \mathbbm{1}_m & = & a \\
\text{diag}(v) K^\trans \text{diag}(u) \mathbbm{1}_n & = & b
\end{array} \right. $$
$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
\text{diag}(u) K v & = & a \\
\text{diag}(v) K^\trans u & = & b
\end{array} \right. $$
Then, we introduce $\odot$ the coefficient-wise product:
$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
u \odot K v & = & a \\
v \odot K^\trans u & = & b
\end{array} \right. $$
$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrc}
u & = & a \, / \, K v \\
v & = & b \, / \, K^\trans u
\end{array} \right. $$
Sinkhorn's algorithm is thus the following: \\
\begin{algorithm}
	\caption{Sinkhorn}
	\Repeat{convergence of $u, v$}{
		$u \gets a \, / \, K v$ \;
		$v \gets b \, / \, K^\trans u$ \;
	}
\end{algorithm}

\paragraph{Complexity}
It has been proved that the algorithm converges in linear time. Moreover,
an iteration is done in $\mathcal{O}(mn)$ but the computation can be
parallelized. It is also possible to use grid convolution to obtain a
complexity of $\mathcal{O}(n \log n)$.

\paragraph{Dual}
On peut aussi relever la formulation du dual avec régularisation :
$$ W_\gamma(\mu, \nu) = \max_{\alpha, \beta} \alpha^\trans a + \beta^\trans b - \gamma \left( e^{\alpha / \gamma} \right)^\trans K \left( e^{\beta / \gamma} \right) $$

\paragraph{Lien avec la divergence KL}
We write the kernel:
$$ K_\gamma(i, j) = \exp \left( - \dfrac{d_{i, j}}{\gamma} \right) $$
We obtain a new formulation of the entropic regularization with the KL-divergence:
$$ W_\gamma(\mu, \nu) = \min_{P \in U(a, b)} \gamma KL(P | K_\gamma) $$

@@

Now that we have convinced ourselves of the good mathematical foundations
of Optimal Transport, and of the fact that we can actually compute it, we
can wonder how we can put this tool to great use in the setting of machine
learning.

\subs{Compute the optimal transport of what?}

In the previous section we have seen that we have an algorithm to compute
optimal transport (Sinkhorn's algorithm), and that this algorithm can be
parallelized. We can take advantage of this fact to actually compute optimal
transport among huge databases. One of the application of this is similarity
based retrieval.

@@

The color-histogram of an image is a tool that characterizes the color
distribution of an image. Interestingly enough, it has been noticed that
color-histogram are good tool to describe the similarity of two images. Since
we now have at our disposal a tool to compute the distance between two
distribution probability, we can apply it for histograms (that are really not
that far from probability distribution, so much so that the algorithms and
methods that we described earlier are relevant even without any fondamental
change of setting). For a given image we can thus compute the distance between
its histogram and the histogram of images comming from a bigger database:
picking images that have the most similar histogram (in the sense of the
Wasserstein distance) will therefore allow some form of similarity based
image retrieval.

\begin{center}
	\includegraphics[scale=0.3]{image_histo.png}
    \captionof{figure}{Two images and their respective color-histograms,
    embedded in $\R^2$}
\end{center}

@@

We can apply the same technique as before, but instead of considering
color-histograms, we can try to find some kind of distribution coming from
other kind of documents, for example text documents. Some previous work,
like the $\texttt{word2vec}$ framework allow for embedding of natural
language words into an Euclidean space. A text therefore defines some sort
of distribution of probability over this embedded space. As for images we
can then compare this distribution to the distributions of other texts coming
from a database (still in the sense of Wasserstein distance), and hopefully
retrieve text dealing with similar topic, although there may be some
significant differences in the actual vocabulary of those texts. We take
advantage of the geometry of the distribution of probability in the embedded
text, that tries to ensure that similar distributions are coming from texts
dealing with similar topics.

@@

We can go further than just using the Wasserstein to compute distance between
distributions and use this as the main tool for the machine learning problems
that we considered (we previously emphasized retrieval tasks) by introducing
the Wasserstein distance on well studied machine learning settings and
problems. We thus shed a new light on those problems, and the results are
often fruitful. Some authors have coined the term $\textbf{Wassersteinization}$
for this process.

@@

One way to approach this process is to introduce Wasserstein distance as a
loss or fidelity term in optimization problems. This will motivate the interest
toward the study of not only the very Wasserstein distance, but also of its
derivative. More precisely we can show that we can computationaly evaluate
the derivatives of the Wasserstein distance. Therefore we can deal with it a variety of optimization problems, including machine lerning related ones.

@@

\subs{Wasserstein PCA, Geodesics and Inverse problem}

Since the space of distributions with the OT geometry has negative curvature,
we generalize the PCA (ie. projecting on one dimension spaces) into Generalized
Principal Geodesics. We then want to explain stuff out of what we found;
eg. affaire Bonneel vs nounours

>>>>

Optimal Transport theory aims to define a geometry along with a notion of
distance on the set of probability measures over a space.

@@

The history of optimal transport formulations can be traced back to Gaspard
Monge's \emph{Mémoire sur la théorie des déblais et des remblais (1781)},
studying how to move masses of sand or soil from one place to another. The
main motivation of this \emph{Mémoire} being that we can assign a cost to
the act of moving a massic particle from one place to another, and it is
of particular interest to seek for the most optimal way of moving a global
quantity of mass from one place to another.

@@

\paragraph{Formalism of Monge's problem}

We work in a probability space $\Omega$. We consider two probability measures
$\mu$ and $\nu$ on $\mathcal{P}(\Omega)$. We also consider a cost function
$c : \Omega \times \Omega \rightarrow \R$. This function can for instance
be the Euclidean distance. Monge's problem is therefore the following

$$ \min_{T: \Omega \to \Omega} \int_{\Omega} c \left( x, T(x) \right) \mu(x) $$
\vspace{-2mm}
$$ \text{s.t.} \quad \forall A \in \mathcal{P}(\Omega), \, \nu(A) = \mu \left( T^{-1}(A) \right) $$

The function $T$ is called a transport map. It carries the information
regarding the displacement of each elementary unit of probability: the
mass located in $x$ is moved toward $T(x)$. The function $c$ thus allows
to quantify the cost of moving one elementary unit of mass. We want to
minimize the average distance travelled by mass units during the optimal
transport. The condition under which the optimization is performed ensures
that all the mass of $\mu$ has been transported toward the distribution $\nu$.

@@

The issue with such a formulation of an optimal transport problem is that
it does not handle discrete probability measures. Indeed, if $\mu$ has a
Dirac impulse bigger than every Dirac impulse of $\nu$, then the transport
map cannot be build: we are only performing displacements of mass, and not
actual repartition of mass.

@@

Kantorovich came up  with an alternative formulation for the optimal
transport problem, aiming to handle discrete distributions.

@@

\paragraph{Change of transport plan}
The idea of Kantorovich is to change the function $T : \Omega \rightarrow
\Omega$ into a probability $P$ on the product space $\mathcal{P}(\Omega \times
\Omega)$. On the above example, Monge's formulation forbids to split the
masse of size 150 into two separate masses. With Kantorovich's formulation,
this operation becomes accessible: the mass of $150$ is splitted into a mass
of $90$ and a mass of $60$, both sent to different places. In the discrete
case, we can represent what happens through a matrix.

@@

If we denote $D$ the distance matrix of the problem, we obtain the following
formulation in the discrete case:

@@

The two conditions encode the fact that we aim to move all the mass of $\mu$
toward all the mass of $\nu$.

@@

In the continuous setting, we introduce the following set of joint
probabilities, which encodes the analog of the two conditions in the discrete
case, making sure that $\mu$ has been fully transported to $\nu$.

$$ \Pi(\mu, \nu) = \left\{ P \in \mathcal{P}(\Omega \times \Omega) \mid \forall A, B \in \Omega, \, P(A \times \Omega) = \mu(A), \, P(\Omega \times B) = \nu(B) \right\} $$

We can rephrase the definition of $\Pi(\mu, \nu)$ as being the set of
probability measures on $\Omega \times \Omega)$ with marginals $\mu$ and $\nu$.

We consider once again a cost function $c$ defined as in the Monge's
problem. Kantorovich's formulation now becomes the following optimization
problem, where $X \sim \mu$ and $Y \sim \nu$:

$$ \inf_{P \in \Pi(\mu, \nu)} \E_P \left[ c(X, Y) \right] = \int \int c(x, y) P(dx, dy) $$

\paragraph{Existence of solutions} One can wonder whether solutions of
Kantorovich's problem as formulated above exist. A rich litterature exists
on this topic. For instance, in the case of measures on the $\mathbf{R}^n$
space, with absolutely continuous measures $\mu$ and $\nu$ with respect to
the Lebesgue measure, the existence (and uniqueness) of solutions will then
depend on the properties of the cost function. If we consider the cost
function $c_p(x,y) = | x - y |^p$, then we will have the following discussion:

\begin{itemize}
    \item $p > 1$: the strict convexity of $c_p$ ensures that there is a
    unique solution to the Kantorovich problem
    \item $p = 1$: here we can obtain the existence of a minimizer for the
    Kantorovich, but we won't have the unicity of the solution
    \item $p < 1$: in general, there will not exist solutions to the problem
\end{itemize}

More generally the existence of minimizers, as well as their uniqueness
depends heavily on the structure of the space we consider, and the properties
of both the measures and the cost function. A comprehensive study of those
properties can be found in \cite{villani2003topics}.

\subs{Wasserstein distance}

Having formulated this optimization problem allows us to define a distance
between probability measures.

\DEF{
    Given a probability space $\Omega$, a cost function $c : \Omega \times
    \Omega \rightarrow \R$ and two probabilitiy mesures $\mu$ and $\nu$ in
    $\mathcal{P}(\Omega)$, the \textbf{p-Wasserstein-distance} between $\mu$
    and $\nu$ is defined as:
	$$ W_p(\mu, \nu) = \left( \inf_{P \in \Pi(\mu, \nu)} \int \int c(x,
	y)^p P(dx, dy) \right)^{1 / p} $$
	\vspace{-3mm}
}

This notion of distance allows to compute barycenters between several
probability distributions, and obtain results that are way more natural
than barycenters obtained with other norms, like the $l^2$ norm for
instance. The following figure compares an interpolation (\emph{i.e.}
a progressive weighted barycenter of two distributions) obtained under
Wasserstein-distance and a linear interpolation (the two target distributions
being the blue and the red one, and the purple shape being an intermediate
interpolation).

@@

What emerges from those two interpolations is that the Wasserstein distance
seems particularily well suited to manipulate distributions, as it appears
to take into account the specificities (like modes, shape, etc) of both
the source and the target distribution, and the resulting barycenter looks
like a distribution combining those into one distribution, while the $l^2$
barycenter does not merge those specificities into one distribution but
rather renormalize the source and target distributions.

We can extend this idea to the computation of the barycenter of an arbitrary
number of distributions. Formally, the computing the barycenter $\mu$ of $N$
distributions $\nu_i$ weighted by the $\lambda_i$ boils down to the following
minimization problem:

$$ \min_{\mu \in \mathcal{P}(\Omega)} \sum_{i=1}^N \lambda_i W_p^p(\mu,
\nu_i) $$

Now we can observe what the barycenter looks like with $4$ distributions
(seen as images) on $\mathbf{R}^2$ in the \figurename~\ref{shp1}.

\begin{figure}[h]
	\centering \includegraphics[scale=0.4]{wasserstein_bar.png}
	\captionsetup{justification=centering} \caption{Each image contains
	several barycenters of the four shapes that are in the corners}
	\label{shp1}
\end{figure}

@@

Now that we have a convenient formulation of the optimal transport problem,
we can study how to effectively compute minimizers of the optimal transport
problem. For this we will introduce several new formulation, more adapted
to effective solving. Sometimes, it is worth noting that we are only interested
in computing the Wasserstein distance between two distributions, rather than
the actual transport map.

@@

To effectively solve the optimal transport problem, we embrace the setting
of discrete spaces. In this setting, probability measures can be seen as a
sum of diracs impulses.

$$ \mu = \sum_{i=1}^n a_i \delta_{x_i} \qquad \nu = \sum_{j =1}^m b_j \delta_{y_j} $$

The cost function is a matrix $D \in \R_+^{n \times m}$ such that $D_{ij}
= d(x_i, y_j)^p$. Our joint distribution $P$ is also a matrix of $\R_+^{n
\times m}$ that has to satisfy several constraints that can be expressed
matricially: if we denote by $a$ and $b$ the vector of the coefficients of
the Dirac decompositions of, the set which $P$ has to belong to is:

$$ U(a, b) = \left\{ P \in \R_+^{n \times m}~|~P \mathbf{1}_m = a, \, P^\trans \mathbf{1}_n = b \right\} $$

@@

\paragraph{Dual} We now focus on the dual form of this optimization
problem. For this we introduce the Lagrangian, using the Lagrange multipliers
$\alpha$ and $\beta$

$$
    L(P, \alpha, \beta) = \langle P, D \rangle + \alpha^\trans (a - P \mathbf{1}_m) + \beta^\trans (b - P^\trans \mathbf{1}_n)
$$

As usual in a Lagrangian formulation, the objective function of the dual
problem is:

$$ g(\alpha, \beta) = \min_{P \geqslant 0} L(P, \alpha, \beta) = \min_{P \geqslant 0} \alpha^\trans a + \beta^\trans b + \langle P, D - \alpha \mathbf{1}_m^\trans - \mathbf{1}_n \beta^\trans \rangle $$

If the matrix $D - \alpha \mathbf{1}_m^\trans - \mathbf{1}_n \beta^\trans$
has a negative coefficient, then by having the corresponding coefficient
of $P$ grow toward $+\infty$ we obtain a minimum of $-\infty$. But
if all the coefficients are non negative, then  $\langle P, D - \alpha
\mathbf{1}_m^\trans - \mathbf{1}_n \beta^\trans \rangle$ is non negative
and we just have to take $P=0$ to obtain a non zero value. We therefore obtain:

$$ g(\alpha, \beta) = \left\{ \begin{array}{ll}
\alpha^\trans a + \beta^\trans b & \text{si } D - \alpha \mathbf{1}_m^\trans - \mathbf{1}_n \beta^\trans \geqslant 0 \\
-\infty & \text{sinon}
\end{array}\right. $$

@@

We have reduced our initial problem into a linear programming
formulation. Using a minimum flow solver, we can solve this linear optimization
problem in $\mathcal{O}(n^3 \log (n))$. But the solution is not stable,
as illustrated on the following figure. If we slightly modify our distance
$D$ or if we change the source and destination measures, the value of $P$
can drastically change, since $P$ is a vertex of our simplex. One solution
is to regularize the problem to fall back on optimizing in a set $U_\alpha(a,
b)$ strictly convex. This set is strictly contained in $U(a, b)$ and the
optimal solution of this new problem will not necessarily be optimal for the
initial problem, but will remain valid.

We will introduce in the next section a regularization of the original
problem, that will us to exhibit an iterative algorithm that converges toward
a meaningful approximation of the minimizer, as well as mainly relying on
matrix-product kind of linear algebrea and thus potentially implemented
on GPUs.

\begin{center}
	\begin{tikzpicture}[scale=20, >={latex}]
		\clip (0.2, -0.01) rectangle (0.51, 0.2);
		\draw[blue, thick] (0.1420, 0.1407) -- (0.5970, -0.1407);
		\draw[black, ->, thick] (0.2038, 0.1216) -- (0.2507, 0.1975);
		\fill[brown!40] (0.4503, 0.1444) -- (0.2251, 0.1444) -- (0.3695, 0.0000) -- (0.4503, -0.0000);
		\fill[greenTikz!60] (0.3827, 0.0946) -- (0.3827, 0.1104) -- (0.3804, 0.1148) -- (0.3759, 0.1206) -- (0.3692, 0.1263) -- (0.3602, 0.1307) -- (0.3556, 0.1321) -- (0.3489, 0.1336) -- (0.3264, 0.1336) -- (0.3219, 0.1321) -- (0.3151, 0.1278) -- (0.3129, 0.1234) -- (0.3129, 0.1191) -- (0.3151, 0.1133) -- (0.3241, 0.1018) -- (0.3444, 0.0888) -- (0.3534, 0.0859) -- (0.3714, 0.0859) -- (0.3804, 0.0917);
		\node[blue] at (0.3444, 0.0888) {$\bullet$};
		\node[blue, below=2] at (0.3444, 0.0888) {$P_\alpha$};
		\node[greenTikz] at (0.3489, 0.1119) {$\bullet$};
		\node[red] at (0.3695, 0.0000) {$\bullet$};
		\node[red, above=2] at (0.3695, 0) {$P$};
		\node at (0.482, 0.08) {$U(a, b)$};
		\node at (0.413, 0.12) {$U_\alpha(a, b)$};
		\node at (0.213, 0.165) {$D$};
	\end{tikzpicture}
\end{center}

\subs{Entropic Regularization}

The regularization that we highlight consists in solving the optimal
transport problem, while maximizing the entropy of the joint-probability
$P$. We therefore exhibit a new parameter $\gamma$ that parametrizes the
regularization and we obtain the following distance.

@@

\paragraph{Remark}
Recall that we can consider two random variables $X$ and $Y$ such that $X \sim
\mu$, $Y \sim \nu$ and consider their joint distribution $P$ such that $(X,Y)
\sim P$. The entropy of $P$ is maximal when $X$ and $Y$ are independant,
\emph{i.e.} when $P$ factorizes as $P = ab^\trans$ and this maximum entropy
is given by $H(\mu) + H(\nu)$. The regularization therefore produces a
matrix whose positive values are more distributed. In other words, we favor
randomness of the transport map against determinism. The intuition behind
this regularization comes from the observation that actual trafic patterns
in networks are sparser than the theoretical solution obtained through
a minimizer of the Kantorovich problem.

@@

By a strict convexity argument, there exist a unique matrix $P_\gamma$
minimizing the distance:
$$ P_\gamma = \argmin_{P \in U(a, b)} \langle P, D \rangle - \gamma H(P) $$

@@

We write the Laplacian of our new optimization problem.

$$ L(P, \alpha, \beta) = \sum_{ij} \left( P_{ij} D_{ij} + \gamma P_{ij} \left( \log_2 P_{ij} - 1 \right) \right) + \alpha^\trans (a - P \mathbf{1}_m) + \beta^\trans (b - P^\trans \mathbf{1}_n) $$

We compute the partial derivative with respect to each $P_{ij}$
$$ \dfrac{\partial L}{\partial P_{ij}} = D_{ij} + \gamma \log_2 P_{ij} - \alpha_i - \beta_j $$

The dual problem is the maximization of the minimum of the Laplacian with
respect to $P$. We must look for $P$ such that the partial derivative is
equal to zero:

$$ \dfrac{\partial L}{\partial P_{ij}} = 0 \Rightarrow P_{ij} = e^{\frac{\alpha_i}{\gamma}} e^{-\frac{D_{ij}}{\gamma}}  e^{\frac{\beta_j}{\gamma}} = u_i K_{ij} v_j $$

\findem

\paragraph{Sinkhorn}
We can derive an iterative method to find this matrix $P_\gamma$. We use
the condition $P_\gamma \in U(a,b)$:

$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
\text{diag}(u) K \text{diag}(v) \mathbbm{1}_m & = & a \\
\text{diag}(v) K^\trans \text{diag}(u) \mathbbm{1}_n & = & b
\end{array} \right. $$

$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
\text{diag}(u) K v & = & a \\
\text{diag}(v) K^\trans u & = & b
\end{array} \right. $$

Then, by introducing $\odot$ the coefficient-wise product, we can rewrite
this as

$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
u \odot K v & = & a \\
v \odot K^\trans u & = & b
\end{array} \right. $$

We end up considering a problem which was already studied in the numerical
analysis community, and known as the \emph{matrix scaling problem}. From
this formulation, rewritten as below using the coefficient-wise division
$\oslash$, we can obtain an iterative algorithm performing Sinkhorn's updates.

$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrc}
u & = & a \oslash K v \\
v & = & b \oslash K^\trans u
\end{array} \right. $$

First we modify $u$ such that it satisfies the first equation, and we then
use this value to modify the value of $v$ in order for it to satisfy the
second. And we keep performing those updates until convergence. Sinkhorn's
algorithm is thus the following: \\

\begin{algorithm}
    \caption{Sinkhorn}
    \Repeat{convergence of $u, v$}{
        $u \gets a \oslash K v$ \;
        $v \gets b \oslash K^\trans u$ \;
    }
\end{algorithm}

\paragraph{Complexity}
It has been proved that the algorithm is a linear convergence scheme. Moreover,
an iteration is naively done in $\mathcal{O}(mn)$ but the computation can
easily be parallelized, and especially performed on GPUs, the bottleneck of one
update being the two matrix products $Kv$ and $K^\trans u$. It is also
possible to use grid convolution to obtain a complexity of $\mathcal{O}(n
\log n)$ for one update. More practical precise and exhaustive details can
be found in Gabriel Peyré and Marco Cuturi's \emph{Computational Optimal
Transport} \cite{Peyr2018ComputationalOT}.

\paragraph{Dual}

One can also notice the formulation of the dual thanks to the Proposition 1.
$$ W_\gamma(\mu, \nu) = \max_{\alpha, \beta} \alpha^\trans a + \beta^\trans b - \gamma \left( e^{\alpha / \gamma} \right)^\trans K \left( e^{\beta / \gamma} \right) $$

\paragraph{Link with the KL divergence}
Writing the kernel:

$$ K_\gamma(i, j) = \exp \left( - \dfrac{d_{i, j}}{\gamma} \right) $$

We obtain a new formulation of the entropic regularized problem using the
KL-divergence

$$ W_\gamma(\mu, \nu) = \min_{P \in U(a, b)} \gamma KL(P | K_\gamma) $$

@@

Now that we have convinced ourselves of the good mathematical foundations
of Optimal Transport, and of the fact that we can actually compute it to
some extent, we can wonder how we can put this tool to use in the setting
of machine learning.

\subs{What do we want to optimally transport ?}

In the previous section we have seen that we have at our disposal an
algorithm to compute optimal transport maps (Sinkhorn's algorithm), and that
this algorithm can be parallelized. We can take advantage of this fact to
actually compute optimal transport among huge databases. Since computing
optimal transport maps implies that we can compute a meaningful distance
between two objects, as long as they can be interpreted as distributions,
one of the application of this is similarity based retrieval.

@@

The color-histogram of an image is a tool that characterizes the
distribution of colors of an image. Interestingly enough, it has been noticed
that color-histograms are a good tool to describe the similarity of two
images. Since we now have a tool to compute the distance between two
distribution probability, we can apply it to histograms (that are essentially
discretized distribution, so much so that the algorithms and methods that
we described earlier are relevant even without any fondamental change of
setting). For a given image we can therefore compute the distance between its
histogram and the histogram of images comming from a bigger database: picking
images that have the most similar histogram (in the sense of the Wasserstein
distance) will therefore allow to perform some form of similarity based
image retrieval.

\begin{center}
	\includegraphics[scale=0.3]{image_histo.png}
    \captionof{figure}{Two images and their respective color-histograms,
    embedded in $\R^2$ (Image from Marco Cuturi's Primer on Optimal Transport at
    NIPS 2017)}
\end{center}

@@

We can apply the same technique as before, but instead of considering
color-histograms, we can try to find some kind of distribution coming from
other kind of documents, for example text documents. Some previous work,
such as the $\texttt{word2vec}$ framework allows for the embedding of
natural language words into an Euclidean space. A text (seen as a collection
of words) therefore somehow defines a probability distribution over this
embedded space. As for images we can then compare this distribution to
the distributions of other texts coming from a database (still in the
sense of Wasserstein distance), and hopefully retrieve text dealing with
similar topics, although there may be some significant differences in the
actual vocabulary of those texts. We can take advantage of the geometry of the
distribution of probability in the embedded text space, that tries to ensure
that similar distributions are coming from texts dealing with similar topics.

@@

We can go further than just using the Wasserstein to compute distance between
distributions and use this notion as distance as a main tool for already
existing machine learning problems (we previously emphasized retrieval
tasks for instance) by introducing the Wasserstein distance on well studied
machine learning settings and problems. We therefore shed a new light on those
problems, and the results are often fruitful. Some authors have coined the
term $\textbf{Wassersteinization}$ to describe this process.

@@

One way to approach Wassersteinization is to introduce Wasserstein distance
as a loss or fidelity term in optimization problems. This will motivate
the interest toward not only the very Wasserstein distance, but also its
derivative. More precisely one can show that we can computationaly evaluate
the derivatives of the Wasserstein distance. Therefore we can deal with it
in a variety of optimization problems, including machine lerning related ones.

@@

That is, we want to propagate the distribution known at certain vertices
along the graph, with a "Wasserstein manner". It is a formulation of several
semi-suppervised learning problems, where we only have at our disposal the
structure of the graph, and information a small portion of the vertices. The
distribution that we are handling here can for instance be probability
distribution of belonging to each class, in a classification problem (which
leads us to solving the "stereotypical" semi-suppervised problem that is
vertice labelling), but we could consider broader classes of problems with
this formalism.

\subs{Wasserstein PCA and Geodesics}
